---
title: "01_spatial data"
author: "Brenna Kelly"
date: "2024-09-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Set-up

```{r}

pkgs <- c("tmap", "ggplot2", "dplyr")

# installs the packages you don't already have
install.packages(pkgs[which(!pkgs %in% installed.packages()[, 1])])

```

The goal of this lab is to get you comfortable reading, writing, and plotting polygon and point data in R using simple features. This will prepare you for simple mapmaking (Module 2) and spatial analysis (Modules 3 & 4).

```{r}

library(sf)
library(tmap)
library(scales)

```

### `simple features`

Check the contents of the `data/la_ozone` folder. This dataset contains July 1996 ozone measurements for Los Angeles basin monitoring stations.  

```{r}

list.files("data/la_ozone")

```

There are four types of files in the `la_ozone` folder:  
- `dbf`: stores the feature attributes in a tabular format  
- `prj`: contains the projection and coordinate reference system  
- `shp`: contains the geometry  
- `shx`: indexes the geometry  

Read in the shapefile `la_o3.shp` using `st_read()`, then check its `class()`.

```{r}

la_o3 <- st_read("data/la_ozone/la_o3.shp")

class(la_o3)

```

You can inspect the contents like a normal dataframe, but we can also map the geometry using the syntax `plot(df$geometry)`.

```{r}

head(la_o3)

plot(la_o3$geometry, col = "red")

```

Plot the data from a column with the geometry using the syntax `plot(df["col"])`.

```{r}

plot(la_o3["MAXDAY"])

```

Next read in `data/california/ca_counties.shp`, which contains polygons for counties in California, and map its geometry.

```{r}

ca_counties <- st_read("data/california/ca_counties.shp")

plot(ca_counties$geometry)

```

Let's map the LA and California data together using the `add = TRUE` argument.

```{r}

plot(ca_counties$geometry)
plot(la_o3$geometry, col = "red", add = TRUE)

```

### Coordinate Reference Systems

The LA and California data appear to be misaligned, probably due to conflicting *coordinate reference systems ((CRS))CRS)*. We're plotting e're using two-dimensional coordinates, the CRS defines how the Earth is "flattened" for mapping. For our purposes, you'll need to know:  
- The CRS contains the rules for projection and the datum being used (origin point on Earth's surface, unit of coordinates, and orientation).  
- Difference reference systems exist, and they are designed to optimize different features (shape, distance, area). Your CRS choice should match your analysis needs.  
- When using multiple datasets, your systems needs to match.  

You can check the coordinate reference system of a simple feature using `st_crs()`. Compare the CRS of the LA data against that of the California data.

```{r}

st_crs(ca_counties)

st_crs(la_o3)

st_crs(ca_counties) == st_crs(la_o3)

```

To fix this, we need to transform one simple feature to the CRS of the other using `st_transform()`.

```{r}

la_o3 <- st_transform(la_o3, crs = st_crs(ca_counties))

```

Check that the CRS matches, then try plotting both features together again.

```{r}

st_crs(la_o3) == st_crs(ca_counties)

# map together
plot(ca_counties$geometry)
plot(la_o3$geometry, col = "red", add = TRUE)

```

**Optional:** transform the California data to two different coordinate reference systems (crs = 3395) and compare.

```{r}

ca_mercator <- st_transform(ca_counties, crs = 3395)
ca_albers <- st_transform(ca_counties, crs = 4326)
ca_original <- ca_counties

plot(ca_mercator$geometry)
plot(ca_albers$geometry)
plot(ca_original$geometry)

```

### Creating and writing `simple features`  

Often you'll find coordinates in a data set, but it hasn't been converted to a simple feature.

```{r}

spiral <- read.csv("data/spiral/spirals.csv")

head(spiral)

```

We could simply plot the x- and y-coordinates, but without a CRS, we cannot combine this with any other information, and we won't get much further use out of the data.

```{r}

plot(spiral$x, spiral$y)

```

To create a simple feature, we use the `st_as_sf()` function. This requires us to specify the column names for the coordinates (e.g., x and y, longitude and latitude) and the EPSG code for the coordinate reference system (if you don't know what it should be, it's easily google-able).

```{r}

spiral_sf <- st_as_sf(spiral, coords = c("x", "y"),
                      crs = 4326)

```

Now that we have our `spiral_sf`, we can plot the geometry.

```{r}

plot(spiral_sf$geometry)

```

To save an `sf` object, we use the `st_write()` function, similar to `write.csv()`.

```{r}

st_write(spiral_sf, "data/spiral/spiral_sf.shp", append = FALSE)

```

### 

Next read in the shapefile of UK Local Authority Districts (located at `data/uk_pubs/uk.shp`). Check the contents of the dataframe, then plot the geometry.

```{r}

uk <- st_read("data/uk_pubs/uk.shp")

head(uk)

plot(uk$geometry)

```

Map the data using the syntax `plot(df["col"])`.

```{r}

plot(uk["LAD24NM"])

plot(uk["LAT"])

plot(uk["LONG"])

```

Read in the shapefile for pubs in the UK (location: `data/uk_pubs/pubs.shp`), and confirm that the coordinate reference system of `pubs` matches that of `uk`.

```{r}

pubs <- st_read("data/uk_pubs/pubs.shp")

st_crs(pubs) == st_crs(uk)

```

Plot the UK district polygons and pubs points together using the `add = TRUE` argument.

```{r}

plot(uk$geometry)
plot(pubs$geometry, col = "goldenrod", add = TRUE) # col = alpha("gold", 0.1)

# check the contents visually
plot(pubs["region"])

plot(uk$geometry)
plot(pubs["region"], add = TRUE)

```

Inspect the district names (`LAD24NM`) in the UK dataset, and choose one district to subset using `subset(uk, LAD24NM == district_name)`.

```{r}

sort(unique(uk$LAD24NM))

highland <- subset(uk, LAD24NM == "Highland")

```

Plot the UK dataset and your chosen district together.

```{r}

plot(uk$geometry)
plot(highland$geometry, col = "seagreen", add = TRUE)

```

You can filter the pubs to only include those contained within your chosen district. Do this using `st_filter(pubs, district_name)`, then plot the district and pubs together.

```{r}

highland_pubs <- st_filter(pubs, highland)

plot(highland$geometry)
plot(highland_pubs$geometry, col = "goldenrod", add = TRUE)

```


### Resources:  
- Simple features: https://r-spatial.github.io/sf/  
- Coordinate reference systems: https://docs.qgis.org/3.34/en/docs/gentle_gis_introduction/coordinate_reference_systems.html#overview
