---
title: "02_mapmaking"
author: "Brenna Kelly"
date: "2024-09-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Set-up

```{r}

library(sf)
library(tmap)
library(dplyr)
library(units)
library(tufte)
library(ggplot2)
library(stringr)
library(RColorBrewer)

```

The goal of this lab is to make simple but effective maps using the `tmap` package.

### What makes an effective map?  

> "Above all else, show the data."
>
> `r tufte::quote_footer('Edward Tufte')`

Like any visualization, a map should communicate a clear message as an honest representation of the data.

```{r}

uk <- st_read("data/uk_pubs/uk.shp")
pubs <- st_read("data/uk_pubs/pubs.shp")

# a simple (but not very useful) representation
plot(uk$geometry)
plot(pubs$geometry, col = "goldenrod", add = TRUE) # col = alpha("goldenrod", 0.1),

# we can aggregate pubs to the district level
pubs_uk <- st_join(pubs, uk,
                   join = st_intersects) |>
  count(LAD24NM) |>
  rename(pub_count = n)
pubs_uk_poly <- merge(uk, st_drop_geometry(pubs_uk), by = "LAD24NM")

# and map pubs within districts
tm_shape(pubs_uk_poly) +
  tm_polygons(col = "pub_count", style = "cont", palette = "plasma",
              border.col = "white", lwd = 0.2, title = "Pub Count") +
  tm_layout(legend.position = c(0.8, 0.5), frame = FALSE)


# transform the count of pubs to density of pubs
pubs_uk_poly$land_area_sqkm <- st_area(pubs_uk_poly)
pubs_uk_poly$land_area_sqkm <- set_units(pubs_uk_poly$land_area_sqkm, km^2)

pubs_uk_poly$pubs_per_sqkm <- pubs_uk_poly$pub_count / pubs_uk_poly$land_area_sqkm

pubs_uk_poly[which(pubs_uk_poly$pubs_per_sqkm ==
                     max(pubs_uk_poly$pubs_per_sqkm)), ]$LAD24NM

# and map the pub density
tm_shape(pubs_uk_poly) +
  tm_polygons(col = "pubs_per_sqkm", palette = "plasma",
              border.col = "white", lwd = 0.2, title = expression("Pubs / km" ^ 2)) +
  tm_layout(legend.position = c(0.8, 0.5), frame = FALSE)

summary(pubs_uk_poly$pubs_per_sqkm)

ggplot(pubs_uk_poly, aes(x = pubs_per_sqkm)) +
  geom_histogram()

# map quantiles
tm_shape(pubs_uk_poly) +
  tm_polygons(col = "pubs_per_sqkm", style = "quantile", palette = "plasma",
              border.col = "white", lwd = 0.2, title = expression("Pubs / km" ^ 2)) +
  tm_layout(legend.position = c(0.8, 0.5), frame = FALSE)

# calculate deciles
pub_breaks <- drop_units(
  quantile(pubs_uk_poly$pubs_per_sqkm,
           seq(from = 0, to = 1, by = 0.1))
)
# clean labels
break_labels = paste("≤", as.character(round(pub_breaks, 2)[-1] |>
                                         str_pad(width = 4, side = "right", pad = "0")))

# map with deciles
tm_shape(pubs_uk_poly) +
  tm_polygons(col = "pubs_per_sqkm", palette = "plasma", breaks = pub_breaks,
              labels = break_labels, border.col = "white", lwd = 0.2,
              title = expression("Pubs / km" ^ 2)) +
  tm_layout(legend.position = c(0.8, 0.5), frame = FALSE)


```

### `tmap` syntax

We will plot the UK data as `polygons` and the pubs data as `dots`, then plot them together.

```{r}

tm_shape(uk) +
  tm_polygons()

tm_shape(pubs) +
  tm_dots()

tm_shape(uk) +
  tm_polygons() +
  tm_shape(pubs) +
  tm_dots()

```

We'll try changing the `tm_polygon()` arguments `col = ` and `border.col = `. For `tm_dots()`, we'll modify the `col = `, `palette = `, `alpha = `, and `title = ` arguments. In addition, add the following line: `tm_layout(legend.position = c(0.8, 0.5), frame = FALSE)`.

```{r}



```

### The choropleth map

A choropleth represents the geographic pattern of statistical variables using color to signify the value in a given area. Usually this is done using **class intervals**, which we specify using the `style = ` argument. The default class interval for `tmap` is `style = "pretty"`, which chooses class intervals which are likely to be legible. Other options include:
"sd", "equal", "quantile", "kmeans", "hclust", "bclust", "fisher", "log10", and "cont".

Read in the `ComArea_ACS14_f.shp` file (located at `data/comarea/ComArea_ACS14_f.shp`), which contains health and socioeconomic indicators for community areas in Chicago from 2013. Check the distribution of the `BrstCancr` variable.

```{r}



```

Create a choropleth of the breast cancer mortality rate using `palette = "viridis"` and `lwd = 0`.

```{r}



```

Modify your code to use an alternative class interval style. Compare with your neighbor.

```{r}



```

> "At the heart of quantitative reasoning is a single question: Compared to what?"
>
> `r tufte::quote_footer('Edward Tufte')`

Modify your code above:  
- Plot the variable `TeenBirth`  
- Remove the `lwd = 0` argument  
- Set `style = "cont"`  
Name this object **border_default**.

```{r}



```

Now create three more maps:  
- **border_black**: add the argument `border.col = "black"`  
- **border_white**: add the argument `border.col = "white"`  
- **border_none**: add the argument `lwd = 0`  
Compare all four maps using `tmap_arrange()`

```{r}



```

Modify your code with the `lwd = ` argument to create the following maps:
- **lwd_wide**: add the argument `lwd = 2`  
- **lwd_default**: drop the `lwd = ` argument  
- **lwd_non**: add the argument `lwd = 0`  
Compare all three maps using `tmap_arrange()`

```{r}



```


#### Mapping populations

We have plotted the community areas based on their boundaries (`tm_polygons()`), but we can also use `tm_dots(size = "Pop2014)` to plot the the areas according to their population size. Compare the two maps using `tmap_arrange()`.

```{r}



```


[Class intervals documentation](https://www.rdocumentation.org/packages/classInt/versions/0.1-7/topics/classIntervals)
