---
title: "03_spatial regression"
author: "Brenna Kelly"
date: "2024-09-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}

library(tmap)
library(spdep)
library(RColorBrewer)

```


Basic spatial regression workflow:  
- **Inspect** the data  
- Define the **spatial structure** (neighbors) and create the **weight matrix**  
- Check for spatial autocorrelation in the outcome (Moran's I)  
- **Model 0:** Create an aspatial null model  
-- Check for SA in the M0 residuals  
-- Interpret  
- **Model 1:** Create an aspatial model with covariates  
-- Check for SA in the M1 residuals  
-- Interpret  
- **Model 2:** Choose a spatial regression method  
-- Lag vs. error vs. hierarchical  
-- Check for SA in the M2 spatial  
-- Interpret  

#### Inspect the data

Read in the shapefile for Chicago communities (location: `data/comarea/ComArea_ACS14_f.shp`). We are interested in estimating the effect of poverty `Pov200P` on the preterm birth rate `PretBrth`. Plot the preterm birth rate with a color scale in the `RColorBrewer` library (using `display.brewer.all()` to choose one).

```{r}



```

#### Spatial structure  

Does it look like there is spatial autocorrelation in the outcome? To confirm this statistically using Moran's I, we first need to define the **neighbor structure** using `poly2nb()` and create the **spatial weight matrix** using `nb2listw()`.

```{r}



```

Visualize the spatial structure. First extract the coordinates of the community areas using `st_coordinates(st_centroid())`. Then plot the `com` geometry, and add `plot(nb, coords)`.

```{r}



```


#### SA in outcome

Calculate **Moran's I** using `moran.test`, which requires a variable in the data and the weight matrix. We will also add the argument `randomisation = TRUE`. Compare the results against `moran.mc`, which uses a Monte Carlo method to estimate significance with improved precision (add `nsim = 999` and `alternative = "greater"`).

```{r}



```

Using `moran.plot`, we can plot the spatial autocorrelation:

```{r}

moran.plot(com$PretBrth, com_nb_listw, plot = TRUE,
           xlab = "Preterm Birth Rate",
           ylab = "Spatially Lagged Rate")

````

#### Null aspatial model

- Create a linear model of the preterm birth rate.  
- Plot the spatial structure of the residuals.  
- Check for SA in the residuals.  

```{r}



```

#### Aspatial model with covariate  

- Scale the poverty variable for the regression.
- Create a linear model of the preterm birth rate with poverty as a covariate.  
- Check for structure in the residuals.
- Plot the spatial structure of the residuals.  
- Check for SA in the residuals.  

```{r}




```

#### Spatial econometric models  
- **error:** unexplained spatial variation; error which is correlated across spatial units  
- **lag:** influence of neighboring values on unit values; a diffusion process  
- **combination**

#### Spatial error model
- Use `errorsarlm` to create a spatial error model of the preterm birth rate with poverty as a covariate. This only additional argument required is the weight matrix.    
- Plot the spatial structure of the residuals.  
- Check for SA in the residuals.  

```{r}



```

#### Spatial lag model
- Use `lagsarlm` to create a spatial lag model of the preterm birth rate with poverty as a covariate.  
- Plot the spatial structure of the residuals.  
- Check for SA in the residuals.  

```{r}



```

#### Error *and* lag
- Using `sacsarlm`, create a model which uses both lag and error structures.  
- Plot the spatial structure of the residuals.  
- Check for SA in the residuals.  

```{r}



```

#### Spatial Durbin error model  
Within neighbors, correlation between errors and correlation between covariates.
- Using `errorsarlm` with the additional argument `etype = "emixed"`, create a spatial Durbin error model.  
- Plot the spatial structure of the residuals.  
- Check for SA in the residuals.  

```{r}



```

#### Spatial Durbin lag model  
Within neighbors, correlation between outcome (diffusion) and correlation between covariates.  
- Modify the `lagsarlm` ode to include the argument `type = "mixed"`.  
- Perform a visual check and statistical test of spatial structure in the residuals.  

```{r}



```

Which model to interpret?

```{r}



```


