---
title: "12_mapmaking solutions"
author: "Brenna Kelly"
date: "2024-09-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Set-up

```{r}

library(sf)
library(tmap)
library(dplyr)
library(units)
library(tufte)
library(ggplot2)
library(stringr)
library(RColorBrewer)

```

The goal of this lab is to make simple but effective maps using the `tmap` package.

### What makes an effective map?  

> "Above all else, show the data."
>
> `r tufte::quote_footer('Edward Tufte')`

Like any visualization, a map should communicate a clear message as an honest representation of the data.

```{r}

uk <- st_read("data/uk_pubs/uk.shp")
pubs <- st_read("data/uk_pubs/pubs.shp")

# a simple (but not very useful) representation
plot(uk$geometry)
plot(pubs$geometry, col = "goldenrod", add = TRUE) # col = alpha("goldenrod", 0.1),

# we can aggregate pubs to the district level
pubs_uk <- st_join(pubs, uk,
                   join = st_intersects) |>
  count(LAD24NM) |>
  rename(pub_count = n)
pubs_uk_poly <- merge(uk, st_drop_geometry(pubs_uk), by = "LAD24NM")

# and map pubs within districts
tm_shape(pubs_uk_poly) +
  tm_polygons(col = "pub_count", style = "cont", palette = "plasma",
              border.col = "white", lwd = 0.2, title = "Pub Count") +
  tm_layout(legend.position = c(0.8, 0.5), frame = FALSE)


# transform the count of pubs to density of pubs
pubs_uk_poly$land_area_sqkm <- st_area(pubs_uk_poly)
pubs_uk_poly$land_area_sqkm <- set_units(pubs_uk_poly$land_area_sqkm, km^2)

pubs_uk_poly$pubs_per_sqkm <- pubs_uk_poly$pub_count / pubs_uk_poly$land_area_sqkm

pubs_uk_poly[which(pubs_uk_poly$pubs_per_sqkm ==
                     max(pubs_uk_poly$pubs_per_sqkm)), ]$LAD24NM

# and map the pub density
tm_shape(pubs_uk_poly) +
  tm_polygons(col = "pubs_per_sqkm", palette = "plasma",
              border.col = "white", lwd = 0.2, title = expression("Pubs / km" ^ 2)) +
  tm_layout(legend.position = c(0.8, 0.5), frame = FALSE)

summary(pubs_uk_poly$pubs_per_sqkm)

ggplot(pubs_uk_poly, aes(x = pubs_per_sqkm)) +
  geom_histogram()

# map quantiles
tm_shape(pubs_uk_poly) +
  tm_polygons(col = "pubs_per_sqkm", style = "quantile", palette = "plasma",
              border.col = "white", lwd = 0.2, title = expression("Pubs / km" ^ 2)) +
  tm_layout(legend.position = c(0.8, 0.5), frame = FALSE)

# calculate deciles
pub_breaks <- drop_units(
  quantile(pubs_uk_poly$pubs_per_sqkm,
           seq(from = 0, to = 1, by = 0.1))
)
# clean labels
break_labels = paste("≤", as.character(round(pub_breaks, 2)[-1] |>
                                         str_pad(width = 4, side = "right", pad = "0")))

# map with deciles
tm_shape(pubs_uk_poly) +
  tm_polygons(col = "pubs_per_sqkm", palette = "plasma", breaks = pub_breaks,
              labels = break_labels, border.col = "white", lwd = 0.2,
              title = expression("Pubs / km" ^ 2)) +
  tm_layout(legend.position = c(0.8, 0.5), frame = FALSE)


```

### `tmap` syntax

```{r}

tm_shape(uk) +
  tm_polygons()

tm_shape(pubs) +
  tm_dots()

tm_shape(uk) +
  tm_polygons() +
  tm_shape(pubs) +
  tm_dots()

tm_shape(uk) +
  tm_polygons(col = "white", border.col = "black") +
  tm_shape(pubs) +
  tm_dots(col = "region", palette = "Dark2", alpha = 0.5, title = "Region") +
  tm_layout(legend.position = c(0.8, 0.5), frame = FALSE)

```

### The choropleth map

A choropleth represents the geographic pattern of statistical variables using color to signify the value in a given area. Usually this is done using **class intervals**, which we specify using the `style = ` argument. The default class interval for `tmap` is `style = "pretty"`, which chooses class intervals which are likely to be legible. Other options include:
"sd", "equal", "quantile", "kmeans", "hclust", "bclust", "fisher", "log10", and "cont".

Read in the `ComArea_ACS14_f.shp` file (located at `data/comarea/ComArea_ACS14_f.shp`), which contains health and socioeconomic indicators for community areas in Chicago from 2013. Check the distribution of the `BrstCancr` variable.

```{r}

com <- st_read("/Users/brenna/Downloads/comarea/ComArea_ACS14_f.shp")

summary(com$BrstCancr)

ggplot(com, aes(x = BrstCancr)) +
  geom_histogram()

```

```{r}

# calculate quantiles
quantiles <- quantile(com$BrstCancr, probs = c(0, 0.2, 0.4, 0.6, 0.8, 1))

# Assign each observations to a quantile range
com$bc_quantile <- cut(com$BrstCancr, breaks = quantiles,
                       labels = c("0-0.2", "0.2-0.4", "0.4-0.6",
                                  "0.6-0.8", "0.8-1.0"),
                       include.lowest = TRUE)

# what the distribution looks like
ggplot(com, aes(x = BrstCancr, fill = bc_quantile)) +
  geom_histogram(bins = 40) +
  scale_fill_viridis_d() +
  xlab("Teen birth rate, per 1000 women aged 15-19") +
  ylab("Count")

```

Create a choropleth of the breast cancer mortality rate using `palette = "viridis"` and `lwd = 0`.

```{r}

# default
tm_shape(com) +
  tm_polygons(col = "BrstCancr",
              palette = "viridis", lwd = 0)

```

Modify your code to use an alternative class interval style.

```{r}

# all options
int_pretty <- tm_shape(com) +
  tm_polygons(col = "BrstCancr", style = "pretty",
              palette = "viridis", lwd = 0) +
  tm_layout(title = "pretty", frame = FALSE, 
            title.position = c(0.05, 0.6))

int_sd <- tm_shape(com) +
  tm_polygons(col = "BrstCancr", style = "sd",
              palette = "viridis", lwd = 0) +
  tm_layout(title = "sd", frame = FALSE, 
            title.position = c(0.05, 0.6))

int_equal <- tm_shape(com) +
  tm_polygons(col = "BrstCancr", style = "equal",
              palette = "viridis", lwd = 0) +
  tm_layout(title = "equal", frame = FALSE, 
            title.position = c(0.05, 0.6))

int_quantile <- tm_shape(com) +
  tm_polygons(col = "BrstCancr", style = "quantile",
              palette = "viridis", lwd = 0) +
  tm_layout(title = "quantile", frame = FALSE, 
            title.position = c(0.05, 0.6))

int_kmeans <- tm_shape(com) +
  tm_polygons(col = "BrstCancr", style = "kmeans",
              palette = "viridis", lwd = 0) +
  tm_layout(title = "kmeans", frame = FALSE, 
            title.position = c(0.05, 0.6))

int_hclust <- tm_shape(com) +
  tm_polygons(col = "BrstCancr", style = "hclust",
              palette = "viridis", lwd = 0) +
  tm_layout(title = "hclust", frame = FALSE, 
            title.position = c(0.05, 0.6))

int_bclust <- tm_shape(com) +
  tm_polygons(col = "BrstCancr", style = "bclust",
              palette = "viridis", lwd = 0) +
  tm_layout(title = "bclust", frame = FALSE, 
            title.position = c(0.05, 0.6))

int_fisher <- tm_shape(com) +
  tm_polygons(col = "BrstCancr", style = "fisher",
              palette = "viridis", lwd = 0) +
  tm_layout(title = "fisher", frame = FALSE, 
            title.position = c(0.05, 0.6))

int_log10 <- tm_shape(com) +
  tm_polygons(col = "BrstCancr", style = "log10",
              palette = "viridis", lwd = 0) +
  tm_layout(title = "log10", frame = FALSE, 
            title.position = c(0.05, 0.6))

int_cont <- tm_shape(com) +
  tm_polygons(col = "BrstCancr", style = "cont",
              palette = "viridis", lwd = 0) +
  tm_layout(title = "cont", frame = FALSE, 
            title.position = c(0.05, 0.6))

tmap_arrange(int_pretty, int_sd, int_equal,
             int_quantile, int_kmeans, int_hclust,
             int_fisher, int_bclust, #int_log10,
             int_cont, ncol = 3)

tmap_mode(mode = "view")
int_pretty

tmap_mode(mode = "plot")

```

> "At the heart of quantitative reasoning is a single question: Compared to what?"
>
> `r tufte::quote_footer('Edward Tufte')`

Modify your code above:  
- Plot the variable `TeenBirth`  
- Remove the `lwd = 0` argument  
- Set `style = "cont"`  
Name this object **border_default**.

```{r}

border_default <- tm_shape(com) +
  tm_polygons(col = "PretBrth", style = "cont",
              palette = "viridis")

```

Now create three more maps:  
- **border_black**: add the argument `border.col = "black"`  
- **border_white**: add the argument `border.col = "white"`  
- **border_none**: add the argument `lwd = 0`  
Compare all four maps using `tmap_arrange()`

```{r}

border_black <- tm_shape(com) +
  tm_polygons(col = "PretBrth", style = "cont",
              palette = "viridis", border.col = "black")

border_white <- tm_shape(com) +
  tm_polygons(col = "PretBrth", style = "cont",
              palette = "viridis", border.col = "white")

border_none <- tm_shape(com) +
  tm_polygons(col = "PretBrth", style = "cont",
              palette = "viridis", lwd = 0)

tmap_arrange(border_default, border_black,
             border_white, border_none, ncol = 2)

```

Modify your code with the `lwd = ` argument to create the following maps:
- **lwd_wide**: add the argument `lwd = 2`  
- **lwd_default**: drop the `lwd = ` argument  
- **lwd_non**: add the argument `lwd = 0`  
Compare all three maps using `tmap_arrange()`

```{r}

lwd_wide <- tm_shape(com) +
  tm_polygons(col = "PretBrth", style = "cont", lwd = 2,
              palette = "viridis", border.col = "white")

lwd_default <- tm_shape(com) +
  tm_polygons(col = "PretBrth", style = "cont",
              palette = "viridis", border.col = "white")

lwd_none <- tm_shape(com) +
  tm_polygons(col = "PretBrth", style = "cont",
              palette = "viridis", lwd = 0)

tmap_arrange(lwd_wide, lwd_default, lwd_none, ncol = 3)

```


#### Mapping populations

```{r}

polys <- tm_shape(com) +
  tm_polygons(col = "PretBrth", style = "cont",
              palette = "viridis", lwd = 0) +
  tm_layout(frame = FALSE, title.position = c(0.05, 0.6))

dots <- tm_shape(com) +
  tm_dots(col = "PretBrth", style = "cont", 
             scale = 2,
          # show.legend = FALSE,
              size = "Pop2014", palette = "viridis") +
  tm_layout(frame = FALSE, title.position = c(0.05, 0.6))

tmap_arrange(polys, dots)

```




[Class intervals documentation](https://www.rdocumentation.org/packages/classInt/versions/0.1-7/topics/classIntervals)