---
title: "13a_spatial regression explainer"
author: "Brenna Kelly"
date: "2024-09-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(sf)
library(tmap)
library(INLA)
library(spdep)

```

aspatial > spatial

aspatial regression: null model

```{r}

#### set up
scot <- st_read("data/scotlip/scotlip.shp")
# scot <- scot[-c(6, 8, 11), ]

tm_shape(scot) +
  tm_polygons(col = "orange", border.col = "white")

scot$cancer_SIR <- 100000 * scot$CANCER / scot$POP

m1 <- glm(CANCER ~ offset(log(POP)), 
          data = scot, family = poisson(link = "log"))

summary(m1)

#### period prevalence
exp(m1$coefficients["(Intercept)"]) * 100000
100000 * sum(scot$CANCER) / sum(scot$POP)

```

aspatial regression: inference

```{r}

cor.test(scot$AFF, scot$cancer_SIR)

scot$AFF_scaled <- scot$AFF / 10 # increase from 0% to 10% AFF

m1 <- glm(CANCER ~ AFF_scaled + offset(log(POP)), 
          data = scot, family = poisson(link = "log"))

```

interpret

```{r}

# odds ratio for AFF
exp(m1$coefficients["AFF_scaled"])
# statistical significance
summary(m1)

```

investigate spatial structure (visually)

```{r}

## map outcome and predictor
Y <- tm_shape(scot) +
  tm_polygons(col = "cancer_SIR", style = "cont", 
              palette = "plasma", lwd = 0.25, border.col = "white")

X <- tm_shape(scot) +
  tm_polygons(col = "AFF", style = "cont", 
              palette = "plasma", lwd = 0.25, border.col = "white")

# does it look like there's spatial structure in the outcome
tmap_arrange(Y, X)

```

investigate spatial structure (statistically)

```{r}

# is there spatial structure in the outcome
nb <- poly2nb(scot)
scot_listw <- nb2listw(nb, zero.policy = TRUE)
moran.mc(scot$cancer_SIR, scot_listw, nsim = 999)

# spatial structure in the error
scot$error_m1 <- m1$residuals

e <- tm_shape(scot) +
  tm_polygons(col = "error_m1", style = "cont", 
              palette = "Spectral", lwd = 0.25, border.col = "white")

# does it look like there's spatial structure in the error
tmap_arrange(Y, e)

# is there spatial structure in the error
moran.mc(scot$error_m1, scot_listw, nsim = 999)

```

spatial model.

```{r}

# spatial structure
nb <- poly2nb(scot)
nb2INLA("map.adj", nb)
g <- inla.read.graph(filename = "map.adj")

scot$idarea <- 1:nrow(scot)
scot$idarea_2 <- 1:nrow(scot)

# model
m2 <- inla(CANCER ~ AFF_scaled + offset(log(POP)) +
             f(idarea, model = "bym2", graph = g), 
           data = scot,  family = "poisson", 
           control.family = list(link = "log"),
           control.compute = list(dic = TRUE, waic = TRUE))

m2$waic$waic

```

investigate spatial structure (visually)

```{r}

## map error
scot$error_m2 <- m2$residuals$deviance.residuals

e_spatial <- tm_shape(scot) +
  tm_polygons(col = "error_m2", style = "cont", 
              palette = "Spectral", lwd = 0.25, border.col = "white")

# does it look like there's spatial structure in the error
tmap_arrange(e, e_spatial)

```

investigate spatial structure (statistically)

```{r}

# is there spatial structure
moran.mc(scot$error_m2, scot_listw, nsim = 999)

```

interpretation

```{r}

# odds ratio for AFF
exp(m2$summary.fixed$mean[2])
# statistical significance
1 - inla.pmarginal(0.95, m2$marginals.fixed$AFF_scaled)

```

error vs. lag
-  error: error which is correlated across spatial units  
-  lag: influence of neighboring values on unit values  

```{r}
# 
# lm1 <- lm(cancer_SIR ~ AFF, data = scot)
# 
# lmt <- lm.LMtests(lm1, scot_listw, test = c("LMerr","LMlag"))
# # best explanation from RSerr
# 
# summary(lmt)

```



